---
title: "Stat 585 - Working with Date and Time"
author: "Heike Hofmann"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, default-fonts, "tweaks.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
```

## `lubridate` package

- package for working with dates and times

- defines different classes of time: instants, periods, intervals, durations

- defines converter and accessor functions, enables time calculus


Resources:

Cheatsheet for [lubridate](https://rawgit.com/rstudio/cheatsheets/main/lubridate.pdf) package

Chapter [Dates and Times](https://r4ds.had.co.nz/dates-and-times.html) in R for Data Science

---

## Converter functions and instants in time

- time instants: one (absolute) moment in time, e.g. `now()`, `Jan-1-2000`

- easy-to-use converter functions: 
    - date: `ymd`, `mdy`, `dmy`, ...
    - time: `hm`, `hms`, ...
    - date & time: `ymd_hms`, `mdy_hm`, ...
    
- order of letters determines how strings are parsed

- separators are automatically determined, then assumed to be the same

Example:
```{r message=FALSE}
mdy("04-18-2022")
```

---

## Accessor functions


- accessor functions: `year`, `month`, `week`, `wday`, `mday`, `yday`, `hour`, `minute`, ...

- accessor functions can also be used for setting elements of date and time, e.g. `hour(x) <- 12`

Examples:

```{r}
month(now())
wday(now(), label = TRUE)
```

---

## Intervals and Durations

- Intervals have a *start* and an *end* date/time: absolute difference

- Durations are potentially of relative length (months, leap year, leap second, ...)

```{r}
end_date <- now()
# span is interval, years(1) is duration
span <- end_date - years(1)
span
end_date - days(10)
```

---
class: inverse
## Your Turn

How many days is it until your birthday? What day of the week will it be this year?

On which day of the week were you born?

How often is the 13th of a month a Friday? Based on the last ten years, are there some days that are more likely to be on a Friday than others?


---

## Your Turn Solutions

```{r}
birthday <- ymd("2023-05-01") # (this is my daughter's birthday)
birthday - today()

wday(birthday, label = TRUE)

wday(ymd("2014-05-01"), label = TRUE)
```

---

```{r}
dates <- seq.Date(from=ymd("2010-01-01"), to=ymd("2021-12-31"), by = 1)

thirteens <- dates[mday(dates) == 13]
table(wday(thirteens, label=TRUE))
```

---


```{r, fig.height = 4.5}
dframe <- data.frame(dates = dates, 
                     weekdays = wday(dates, label=TRUE), 
                     month_days = mday(dates))
dframe %>% 
  ggplot(aes(x = month_days, fill = weekdays)) + 
  geom_bar()
```

---

```{r}
xtabs(~weekdays+month_days, data = dframe)
```
