---
title: "Stat 585 - profiling code"
author: "Heike Hofmann"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, default-fonts, "tweaks.css"]
    nature:
      beforeInit: "macros.js"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

## Beyond the: let's just get it to work


Signs of possible inefficiencies

- Violation of the language model
    - e.g. R is vector oriented language, i.e.  ```for``` loops should be used sparingly, 
    - e.g. use of ```rbind``` (it is *known* to be very slow)

- Insufficient knowledge of the language. We're all guilty of that at different levels 

- Reading other people's code helps improve our language knowledge!

---

## What do others say?

- **Ken Thompson:** <br> keep the number of indentations down, high number of local variables is suspicious

- **Hadley Wickham:** <br> functions with more than ten lines are suspicious, as is highly repetitive code

- **Generally Good Practices:** <br> structure code using indentation and spacing, write comments, use clear naming convention


<br><br>


- BUT: always test and don't assume
---

## Claim: loops are slow

`x` is vector with numbers 1,2,3,4

Objective: re-code these into weather events: "Sun", "Rain", "Snow", "Hail"

```{r}
x <- sample.int(4, size=1000, replace = TRUE)
y <- vector("character", length=length(x))
for (i in 1:length(x)) {
   y[i] <- switch (x[i], "Sun", "Rain", "Snow", "Hail", default=NA)
}

table(y)
```

---

## Removing for loops (cont'd)

```{r}
# the switch statement on the previous slide is equivalent to 
# the following lines:
for (i in 1:length(x)) {
  if (x[i] == 1) y[i] <- "Sun"
  if (x[i] == 2) y[i] <- "Rain"
  if (x[i] == 3) y[i] <- "Snow"
  if (x[i] == 4) y[i] <- "Hail"
}

# using R's vector system we can reduce this to a single line:
y <- c("Sun", "Rain", "Snow", "Hail")[x]
```

---

## Removing for loops (cont'd)

All of the previous solutions doctor around symptoms

We are still breaking the language model :)

--

```{r}
y <- factor(x, labels =  c("Sun", "Rain", "Snow", "Hail"))

summary(y)
```

But: this is the only variant that returns a factor variable


--

... now we want to compare these four options


---

## Profiling

Code can be profiled at different levels:

1. two (or more) alternatives of code with the same objectives are timed to determined the one which is the fastest (`system.time()` and `proc.time()`)

2. one piece of code is analyzed to find weak/slow spots (e.g. `profvis`, `benchmarkr` or `microbenchmark` )


---

## Profiling process

- Collect information on code performance:

- memory usage

- frequency and duration of function calls

- Goal: optimization of code


- R has an in-built profiler: `Rprof`  Output in `Rprof.out`


---

## `system.time()` or `proc.time()`

- Probably simplest of all measures

- Evaluates the time spent in the whole expression

- doesn't give any clues about what is done at a lower level <br><br>

---

## Claim: 'Everybody knows that' `read_csv` is faster than `read.csv`


```{r message=FALSE}
data(diamonds, package="ggplot2")
tmpfile <- tempfile(fileext = "csv")
write.csv(diamonds, file = tmpfile)

ptm <- proc.time()
t1 <- read.csv(tmpfile)
proc.time() - ptm

ptm <- proc.time()
t2 <- readr::read_csv(file=tmpfile, progress = FALSE)
proc.time() - ptm
```
---

## Use `rbenchmark`

```{r message=FALSE}

library(rbenchmark)

benchmark(
  "read.csv" = {
    read.csv(tmpfile)
    },
  "read_csv" = {
    readr::read_csv(file=tmpfile, progress = FALSE)
  }
)
```
---

## Everybody knows that `read_csv` is faster for large data?

```{r}
data(diamonds, package="ggplot2")
tmpfile <- tempfile(fileext = "csv")
diamonds5 <- rbind(diamonds,diamonds,diamonds,diamonds,diamonds)
write.csv(diamonds5, file = tmpfile, row.names=FALSE)
```
```{r, message=FALSE}
benchmark(
  "read.csv" = {
    read.csv(tmpfile)
    },
  "read_csv" = {
    readr::read_csv(file=tmpfile, progress = FALSE)
  }
)
```

---

## Reading binary files - fast! 

But introduces platform/version dependencies 

```{r message=FALSE}
tmpfile_rds <- gsub("csv", "rds", tmpfile)
saveRDS(diamonds5, tmpfile_rds)
tmpfile_rda <- gsub("csv", "rda", tmpfile)
save(diamonds5, file=tmpfile_rda)


ptm <- proc.time()
t3 <- readRDS(tmpfile_rds)
proc.time() - ptm

ptm <- proc.time()
load(tmpfile_rda)
proc.time() - ptm
```


---
class: inverse
## Your turn: which option is fastest?

```{r message=FALSE}
x <- sample.int(4, 1000, replace=TRUE)
y <- vector("character", length=length(x))

# Option #1
for (i in 1:length(x)) {
   y[i] <- switch (x[i], "Sun", "Rain", "Snow", "Hail", default=NA)
}

# Option #2
for (i in 1:length(x)) {
  if (x[i] == 1) y[i] <- "Sun"
  if (x[i] == 2) y[i] <- "Rain"
  if (x[i] == 3) y[i] <- "Snow"
  if (x[i] == 4) y[i] <- "Hail"
}

# Option #3
y <- c("Sun", "Rain", "Snow", "Hail")[x]

# Option #4
y <- factor(x)
levels(y) <- c("Sun", "Rain", "Snow", "Hail")
```

---

## Your Turn Result

Based on 500 evaluations of samples of size 5000: 

```{r message=FALSE, warning=FALSE, echo = FALSE, fig.height = 6, fig.width = 7}
library(tidyverse)
set.seed(20230419) # remember the rules of reproducibility!

results <- 1:500 %>% purrr::map_df(.f = function(x) {
  x <- sample.int(4, 5000, replace=TRUE)
  y <- vector("character", length=length(x))
  
  ptmbase <- proc.time()
  # Option #1
  for (i in 1:length(x)) {
    y[i] <- switch (x[i], "Sun", "Rain", "Snow", "Hail", default=NA)
  }
  ptm1 <- proc.time()
  
  # Option #2
  for (i in 1:length(x)) {
    if (x[i] == 1) y[i] <- "Sun"
    if (x[i] == 2) y[i] <- "Rain"
    if (x[i] == 3) y[i] <- "Snow"
    if (x[i] == 4) y[i] <- "Hail"
  }
  
  ptm2 <- proc.time()
  
  # Option #3
  y <- c("Sun", "Rain", "Snow", "Hail")[x]

  ptm3 <- proc.time()
  
  # Option #4
  y <- factor(x, labels=c("Sun", "Rain", "Snow", "Hail"))

  
  ptm4 <- proc.time()
  data.frame(value=c((ptm1-ptmbase)["elapsed"], (ptm2-ptm1)["elapsed"], (ptm3-ptm2)["elapsed"], (ptm4-ptm3)["elapsed"]), Option=1:4)
})

results %>% 
  ggplot(aes(x = Option, y = value, fill= factor(Option))) +geom_boxplot() +
  ylab("Elapsed time") 

```

---

```{r}
results %>% group_by(Option) %>% summarise(eval=mean(value))
```

Option 3 is  fastest on average. But why? - Here we can relatively easily answer the question. Generally: more in-depth profiling 

---

## Profiling Code: frequency and duration of function calls


- R has an in-built profiler: `Rprof`  Output in `Rprof.out`

```{r eval=FALSE}
Rprof()

# ....

Rprof(NULL)
```

- call stack is written every 0.02 seconds (interval) into a temporary file

- output file is processed to produce a summary of the usage

- time measurements varies by platform: on Unix-alike time is CPU time of the R process (excludes time waiting for input)

- Note: profiling takes time, too; once the timer goes off, the information is not recorded until the next timing click (probably in the range 1–10 msecs). Can’t go below that.


---

## `profvis` example

`profvis` package provides a wrapper around `Rprof`


```{r eval=FALSE}
library(profvis)

p <- profvis({			
  data(diamonds, package = "ggplot2")			
  plot(price ~ carat, data = diamonds)			
  m <- lm(price ~ carat, data = diamonds)			
  abline(m, col = "red")			
})
```

`profvis` creates interactive response with code and flame plot.

Copy and paste the code into your console to run it.

---
class: inverse
## Your Turn - Part 1

Write a function to find the factors of a positive integer `x`

---

## Discuss approaches


---
class: inverse
## Your Turn - Part 2


Benchmark the approaches to identify the fastest function.


Don't forget about readability and maintainability of code!
